Constructing ordinal outcomes for clinical trials

### 2025-06-06 constructing_ordinal_outcomes.tex
Bunch of work

@@

Ordinal outcomes have become increasingly prominent in clinical
resarch, particularly following the COVID-19 pandemic, when ordinal
scales developed by the WHO and NIAID were used in prominent trials
of drugs repurposed to treat the disease \todo{cite}. They have also
been a popular choice in adaptive platform trials \todo{cite}.

Some advantages of ordinal outcomes are that they can
\begin{enumerate}
  \item Provide more information that binary outcomes, including
    composite outcomes, making study designs more sensitive and precise, and
    therefore more efficient and ethical

@@

The goal of this manuscript is to serve as a guide for clinicians and trialists
in developing ordinal outcomes for their studies. We consider both the outcome
itself (the data recorded on each patient) and effect measures or ``endpoints''
(the metric used to compare distributions of the outcome across patient
groups).

A final goal of this paper is to situate the literature around ``generalized
pairwise comparisons of prioritized outcomes''
\citep{buyseGeneralizedPairwiseComparisons2022}--including win ratio
\citep{pocockWinRatioNew2012} and Desirability of Outcome Ranking (DOOR)
methods \citep{evansDesirabilityOutcomeRanking2015,
ongUnlockingDOORHow2023}--within the broader context of ordinal outcomes and
endpoints.

@@

\subsection{Constructing versus validating}

@@

Three broad ways to combine ordinal outcomes:
\begin{enumerate}
  \item \textbf{Sequential composition} Tie-breaking by $B$
    \emph{within either highest or lowest level of $A$}. Also could
    call this ``overriding by $A$''. A special case of this is
    \emph{overriding events} when $A$ is binary.
  \item \textbf{Hierarchical composition} Complete tie-breaking of $A$ by
    $B$ (``prioritized outcomes''; lexicographic order) within
    \emph{every} level of $A$.

    This is what win methods do.
  \item \textbf{Parallel composition} This is where levels of $A$ are
    combined with levels of $B$ ``in parallel''; high levels of $A$
    are combined with high levels of $B$ and low levels of $A$ with
    low levels of $B$.

    This is not so precisely determined as the other two. We need to
    decided \emph{how} to aggregate.
\end{enumerate}

Suggested notation:
\begin{itemize}
  \item "$\rightarrow$" and "$\leftarrow$" to denote sequential
    composition/overriding,
  \item "$/$" to denote hierarchical composition/complete tiebreaking
    (note that this is an asymmetric operation), and
  \item "$|$" to denote parallel composition
\end{itemize}

@@

Aside: \citet{wittkowskiCombiningSeveralOrdinal2004} consider combining
ordinal variables into partial orders. This makes more sense in the
generalized pairwise comparisons framework.

\begin{figure}
  \includegraphics[width=7in]{passport_ordinal_outcome_sequential.pdf}
  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
  \caption{Ordinal outcome used in the planned PASSPORT adaptive
    platform trial. This illustrates sequential composition/overriding:
    of days hospitalized by maximum pSOFA while hospitalized, and of
  both of these by day 28 status.}
  \label{fig:passport_outcome}
\end{figure}

\begin{figure}
  \includegraphics[width=6in]{rajkumar_et_al_outcome.png}
  \caption{``Angina Symptom Score'' ordinal outocome develped in
    \citep{rajkumarPlaceboControlledTrialPercutaneous2023}. This illustrates
    (1) complete tie-breaking (of units of antianginal medication by
    episodes of angina) and (2) overriding (of angina episodes and
  angina medication by adverse clinical events)}
  \label{fig:rajkumar_et_al_outcome}
\end{figure}

\begin{figure}
  \includegraphics[width=7in]{parallel_composition_bandicoot.pdf}
  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
  \caption{Ordinal outcome used in the planned BANDICOOT adaptive
    platform trial \citep{walkerCodesigningNovelOrdinal2025}.
  Illustrates \textbf{parallel composition} of component outcomes.}
  \label{fig:parallel_bandicoot}
\end{figure}

\subsection{Converting back to numeric scores}

Treat each input factor as
numeric then sum them up. I believe \citet{ongUnlockingDOORHow2023}
has examples?

@@

\subsection{Pairwise comparison methods (win, DOOR) implicitly construct ordinal
outcomes}

@@

\begin{itemize}
  \item Situating methods in their proper context can help
    trialists make informed choices among them
  \item Win methods can make use of
    plots and diagnostics for general ordinal outcomes, for the
    \emph{implicit} ordinal variable
    \begin{itemize}
      \item
        comparative CDF plots for the
      \item ridit plots
        \citep{brossHowUseRidit1958,jansenRiditAnalysisReview1984}
        relative to a reference group, which should allow one to
        display both full eCDF information and also win statistics
        (based on the interpretation of the mean ridit as a win probability)
    \end{itemize}e.g.
  \item Ordinal regression models can use win methods for model
    checking; it is
    also possible to present a model-based win effect measure
    alongside a common
    odds ratio for example. See
    \citep{agrestiOrdinalProbabilityEffect2017} for example, which
    covers both latent variable and predicted probability versions.
  \item The outcomes \emph{implicitly} constructed using
    DOOR and win methods may contain many levels, and may contain
    levels where an
    effect is not so meaningful (e.g. time-to-death when the followup time is
    short). By \emph{explicitly} constructing the associated
    ordinal outcome one
    might be forced to be more aware of this
  \item What is a reasonable win
    analysis of a specific component score? From the traditional
    ordinal endpoint
    school of thought what would make most sense is to \emph{only
      tie-break for the
    lowest and highest levels of that score}, and e.g. plot eCDFs of this new
    variable. Is this what the plots do that break down wins and losses by
    component score?
\end{itemize}

@@

\section{Reasons to combine ordinal outcomes}

\subsection{Providing a more holistic measure of benefit and harm}

\begin{figure}
  \includegraphics[width=5in]{remap_outcome_extra_cropped.jpg}
  \caption{Ordinal outcome used in the REMAP-CAP adaptive
    platform trial
    \citep{theremap-capinvestigatorsInterleukin6ReceptorAntagonists2021}.
    Illustrates overriding a numeric ordinal outcome (organ
    support-free days) by a binary ordinal outcome (death; this is
    ``sequential composition''), which
    would also be an intercurrent event. This is another example of
  sequential composition.}
  \label{fig:remap_cap_override}
\end{figure}

\subsection{Increasing precision by reducing the probability of the
largest category}

Theory of \citep{whiteheadSampleSizeCalculations1993} says that the relative
efficiency (variance) of an unadjusted treatment comparison using an ordinal
versus continuous outcome in the proportional odds model is given by $1 - \sum
\overline p_c^3$ where $\overline p_c$ is the marginal probability (averaged
over control and experimental groups) of outcome category $c$.

Same should apply to the Mann-Whitney-Wilcoxon test, and therefore to
the win probability, and therefore at least approximately to win
difference/win ratio/win odds.

Efficiency is maximized at $1 - 1/C^2$ where $C$ is the total number of
categories when the probabilities are all equal: $\overline p_c = 1/C$.

We have simulation results and some maths to show that the value of $\sum_c
\overline p_c^3$ is \emph{essentially a function of the largest category
probability} $\overline{p_{\max}} = \max \overline{p_c}$. See Figure
\ref{fig:p_max} for an illustration of the maths.

\begin{figure}
  \includegraphics[width=6in]{p_max_controls_efficiency.png}
  \caption{Bounds on the relative efficiency of an ordinal outcome
    compared to its continuous counterpart as a function of the maximum
    category probability $p_\text{max}$. The upper bound (best case
    relative efficiency for a given $p_\text{max}$) is always achieved in the
    limit of infinitely many categories ($C = \infty$). The lower bound
    is always achieved when the number of categories is as small as
  possible for a given $p_\text{max}$.}
  \label{fig:p_max}
\end{figure}

Thus simple advice is: for statistical efficiency, try to make the proportion
of patients in the largest category small.

However a converse is: if the proportion of patients in the largest
category is
small, not much is gained by tie-breaking.

\subsection{Augmenting a state outcome with longitudinal
information}

Distinguish between ``state'' and ``trajectory'' variables. State
variables summarize patient outcomes at a particular moment in time.
Trajectory variables, including times-to-event, summarize patient
state over a timecourse.

Trajectory and state variables can be incorporated into an ordinal outcome. For
example (from PASSPORT) ``worst pSOFA while hospitalized until day 28'' and
``length of hospitalization'', which are longitudinal variables (summary of a
trajectory), are combined with the day 28 state variable ``death''.

Incorporating trajectory variables into an ordinal outcome may seem unnatural.
But arguably most time-to-event analyses are in fact analyses of an
ordinal outcome. The Cox model for time-to-event data treats time as ordinal:
because (1) the baseline hazard is not modelled, (2) the model itself is
invariant to monotone rescalings of the time axis, and (3) the (partial)
likelihood (ignoring ties) depends only on the ranks of the data (i.e., is a
marginal likelihood). The Cox model is essentially an ordinal
stopping ratio model with complementary log-log link. (Perhaps a
difference is censoring.)

@@

\subsection{Coarsening to avoid directly comparing levels}

@@

(It isn't really a problem for model fitting with modern software.)

>>>>

\begin{abstract}
  Ordinal outcomes or endpoints are an increasingly popular choice in
  clinical trials.
  They have been recently popular in the study of COVID-19 and in
  adaptive platform
  trials.
  This paper is a guide to techniques for constructing ordinal
  outcomes, aimed at clinicians and their statistical collaborators.
  We categorize and illustrate ways combining existing binary,
  ordinal, or continuous
  outcomes through three broad mechanisms: sequential composition
  (overriding), hierarchical composition (tie-breaking), and parallel
  composition.
  We discuss how this fine-graining impacts validity and efficiency, and also
  suggest reasons why coarser outcomes might be preferred.
  We also situate ``generalized pairwise comparisons'' and
  ``prioritized outcomes'' approaches, like the win
  ratio and desirability of outcome
  ranking (DOOR) methodologies, in the broader context of ordinal
  outcomes, suggesting that methods and approaches from each field
  might be useful in the other.
\end{abstract}

@@

Ordinal outcomes (or ``endpoints'') have become increasingly
prominent in clinical
resarch, particularly following the COVID-19 pandemic, when ordinal
scales developed by the WHO and NIAID were used in prominent trials
of drugs repurposed to treat the disease \todo{cite}. They have also
been a popular choice in adaptive platform trials \todo{cite}.

Ordinal outcomes rank patient outcomes from better to worse, but do
not compare \emph{differences} in patient outcomes. Thus binary
outcomes are ordinal, and so are continuous outcomes (though
``treating them ordinally'' means ignoring numeric values).

Some advantages of ordinal outcomes are that they can
\begin{enumerate}
  \item Provide more information than binary outcomes, making study
    designs more sensitive and precise, and
    therefore more efficient and ethical

@@

The goal of this manuscript is to serve as a guide for clinicians and trialists
in developing ordinal outcomes for their studies.

It is important to distinguish ``outcomes'' and ``effect measures'' -- both
sometimes referred to under the banner of ``endpoints''. Here we
focus on outcomes; we discuss endpoints in a separate paper \todo{cite}.

Another goal is to situate the literature around ``generalized
pairwise comparisons of prioritized outcomes''
\citep{buyseGeneralizedPairwiseComparisons2022}--including win ratio
\citep{pocockWinRatioNew2012} and Desirability of Outcome Ranking (DOOR)
methods \citep{evansDesirabilityOutcomeRanking2015,
ongUnlockingDOORHow2023}--within the broader context of ordinal outcomes and
endpoints.

@@

\subsection{The iterative process of constructing and validating}

\section{Reasons to combine (or fine-grain) ordinal outcomes}

\subsection{Capturing a broader spectrum of benefit and harm}

Example: PASSPORT outcome (Figure \ref{fig:passport_outcome}),
designed to capture benefit and harm across sepsis patients with a
range of different disease severities.

\begin{figure}
  \includegraphics[width=5in]{remap_outcome_extra_cropped.jpg}
  \caption{\textbf{Overriding a continuous outcome, illustrated by
    REMAP-CAP}. Outcome from the REMAP-CAP adaptive
    platform trial
    \citep{theremap-capinvestigatorsInterleukin6ReceptorAntagonists2021}.
    Illustrates overriding a numeric ordinal outcome (organ
    support-free days) by a binary ordinal outcome (death; this is
    ``sequential composition''), which
    would also be an intercurrent event. This is another example of
  sequential composition.}
  \label{fig:remap_cap_override}
\end{figure}

\subsection{Increasing efficiency by shrinking the largest category}

Theory of \citep{whiteheadSampleSizeCalculations1993} says that the relative
efficiency (variance) of an unadjusted treatment comparison using an ordinal
versus continuous outcome in the proportional odds model is given by $1 - \sum
\overline p_c^3$ where $\overline p_c$ is the marginal probability (averaged
over control and experimental groups) of outcome category $c$.
Efficiency is maximized at $1 - 1/C^2$ where $C$ is the total number of
categories when the probabilities are all equal: $\overline p_c = 1/C$.

The results are based on estimating a common odds ratio, but in large
enough samples are equivalent to the Mann-Whitney-Wilcoxon test, and
therefore to testing the various ``win nulls'' (win probability is
  0.5, win difference
is 0, win ratio is 1).

We have simulation results and some maths to show that the value of $\sum_c
\overline p_c^3$ -- and therefore the (relative) efficiency of the
ordinal outcome -- is essentially a function of the largest category
probability $\overline{p}_{\max} = \max \overline{p}_c$. See Figure
\ref{fig:p_max} for an illustration of the maths.

\begin{figure}
  \includegraphics[width=6in]{p_max_controls_efficiency.png}
  \caption{\textbf{Efficiency gains are controlled by the probability
    of the largest category.} Bounds on the relative efficiency of an
    ordinal outcome
    compared to its continuous counterpart as a function of the maximum
    category probability $p_\text{max}$. Results are based on
    estimating a common odds ratio or testing stochastic ordering
    using the Mann-Whitney-Wilcoxon test. The upper bound (best case
    relative efficiency for a given $p_\text{max}$) is always achieved in the
    limit of infinitely many categories ($C = \infty$). The lower bound
    is always achieved when the number of categories is as small as
    possible for a given $p_\text{max}$; in particular when
  $p_\text{max} \geq 0.5$ the worst case occurs with 2 categories.}
  \label{fig:p_max}
\end{figure}

Thus simple advice is: for statistical efficiency, try to make the proportion
of patients in the largest category small. However a converse is: if the
proportion of patients in the largest category is small, not much is gained by
fine-graining; a coarser outcome might be easier to communicate.

\subsection{Augmenting a state or time-to-event outcome with
longitudinal information}

Distinguish between ``state'' and ``trajectory'' outcomes. State
variables summarize patient outcomes at a particular moment in time.
Trajectory variables, including times-to-event, summarize patient
state over a timecourse. (In practice ``state'' variables are
actually summaries over an interval like ``day 28 post-randomization''.)

Trajectory and state variables can be incorporated into an ordinal outcome. For
example (from PASSPORT; Figure \ref{fig:passport_outcome}) ``worst pSOFA while
hospitalized until day 28'' and ``length of hospitalization'', which are
longitudinal variables (summary of a trajectory), are combined with the day 28
state variable ``death''. The REMAP-CAP outcome is similar (Figure
\ref{fig:remap_cap_override}).

Most time-to-event analyses are in fact analyses of an ordinal outcome. The Cox
model for time-to-event data treats time as ordinal: because the baseline
hazard is not modelled, the model itself is invariant to monotone rescalings of
the time axis, and the (partial) likelihood (ignoring ties) depends only on the
ranks of the data (i.e., is a marginal likelihood). The Cox model is
essentially an ordinal stopping ratio model with complementary log-log link.
(Plus handling right-censoring.)

An important concern when incorporating longitudinal information is
to avoid including nonclinical effects of treatment/exposure in the
outcome; \citet{ongUnlockingDOORHow2023} discuss/criticize RADAR
methods from this perspective.

Win methods were inaugurated in
\citet{finkelsteinCombiningMortalityLongitudinal1999} with the
express motivation of combining longitudinal and time-to-event information.

@@

Three broad ways to combine ordinal outcomes:
\begin{enumerate}
  \item \textbf{Sequential composition} Tie-breaking by $B$
    \emph{within either highest or lowest level of $A$}. Also could
    call this ``overriding by $A$''. A special case of this is
    \emph{overriding events} when $A$ is binary.

  \item \textbf{Hierarchical composition} Complete tie-breaking of $A$ by
    $B$ (``prioritized outcomes''; lexicographic ordering) within
    \emph{every} level of $A$.

    This is what win methods do.

    Sequential composition is complete-tie breaking followed by coarsening.

  \item \textbf{Parallel composition} This is where levels of $A$ are
    combined with levels of $B$ ``in parallel''; high levels of $A$
    are combined with high levels of $B$ and low levels of $A$ with
    low levels of $B$.

    This is not so precisely determined as the other two. We need to
    decided \emph{how} to aggregate. Some ways are
    \begin{itemize}
      \item summing up numeric
        scores
      \item counting different kinds of events
      \item taking the worst level across ordinal outcomes considered
        in parallel (as in BANDICOOT; Figure \ref{fig:parallel_bandicoot}).
    \end{itemize}
\end{enumerate}

Suggested notation:
\begin{itemize}
  \item "$\rightarrow$" and "$\leftarrow$" to denote sequential
    composition/overriding,
  \item "$/$" to denote hierarchical composition/tiebreaking
    (note that this is an asymmetric operation), and
  \item "$|$" to denote parallel composition
\end{itemize}

@@

Aside: \citet{wittkowskiCombiningSeveralOrdinal2004} consider combining
ordinal variables into \emph{partial} orders. Beyond the scope of this paper.

\begin{figure}
  \includegraphics[width=7in]{passport_ordinal_outcome_sequential.pdf}
  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
  \caption{\textbf{Sequential composition (overriding) illustrated
    with the PASSPORT outcome} Ordinal outcome used in the planned
    PASSPORT adaptive
    platform trial. This illustrates sequential composition/overriding:
    of days hospitalized by maximum pSOFA while hospitalized, and of
  both of these by day 28 status.}
  \label{fig:passport_outcome}
\end{figure}

\begin{figure}
  \includegraphics[width=6in]{rajkumar_et_al_outcome.png}
  \caption{\textbf{Hierarchical composition (complete tie-breaking)
      illustrated with the
    the ORBITA-2 outcome.} ``Angina
    Symptom Score'' ordinal outocome develped in
    \citep{rajkumarPlaceboControlledTrialPercutaneous2023}. This illustrates
    (1) hierarchical composition or complete tie-breaking (of units
      of antianginal medication by
    episodes of angina) and also (2) overriding (of angina episodes and
  angina medication by adverse clinical events)}
  \label{fig:rajkumar_et_al_outcome}
\end{figure}

\begin{figure} \includegraphics[width=7in]{parallel_composition_bandicoot.pdf}
  % \includegraphics[width=1in]{passport_ordinal_outcome_colorbar.png}
  \caption{\textbf{Parallel composition illustrated with the BANDICOOT outcome}.
    Ordinal outcome used in the planned BANDICOOT adaptive platform trial
    \citep{walkerCodesigningNovelOrdinal2025}. Illustrates \textbf{parallel
  composition} of component outcomes.} \label{fig:parallel_bandicoot}
\end{figure}

\subsubsection{Parallel composition by combining numeric scores}

One option: convert each input factor to a numeric value then sum them up.

A specific example of this, described e.g. by \citet{ongUnlockingDOORHow2023} in
the infectious disease context, is counting events of a particular
type (corresponds to converting the presence/absence of the event to
a 1/0 binary variable).

@@

\subsection{Pairwise comparison methods (win and DOOR) implicitly construct
ordinal outcomes}

@@

\begin{itemize}
  \item Situating methods in their proper context can help
    trialists make informed choices among them
  \item Win methods can make use of
    plots and diagnostics for general ordinal outcomes, for the \emph{implicit}
    ordinal variable
    \begin{itemize}
      \item comparative CDF or ``survival''
        (complementary CDF) plots for assessing stochastic
        monotonicity \todo{this
          seems not done routinely? perhaps because of the difficulty
        of incorporating censoring?}
      \item ridit plots
        \citep{brossHowUseRidit1958,jansenRiditAnalysisReview1984} relative to a
        reference group, which should allow one to display both full
        eCDF information
        and also win statistics (based on the interpretation of the
          mean ridit as a win
        probability)
    \end{itemize}e.g.
  \item Ordinal regression models can use win
    methods for model checking; it is also possible to present a model-based win
    effect measure alongside a common odds ratio for example. See
    \citep{agrestiOrdinalProbabilityEffect2017} for example, which covers both
    latent variable and predicted probability versions.
  \item The outcomes
    \emph{implicitly} constructed using DOOR and win methods may contain many
    levels, and may contain levels where an effect is not so meaningful (e.g.
    time-to-death when the followup time is short). By \emph{explicitly}
    constructing the associated ordinal outcome one might be forced to be more
    aware of this
  \item What is a reasonable win analysis of a specific component
    score? From the traditional ordinal endpoint school of thought
    what would make
    most sense is to \emph{only tie-break for the lowest and highest
      levels of that
    score}, and e.g. plot eCDFs of this new variable. Is this what the plots do
    that break down wins and losses by component score?
\end{itemize}

@@

@@

\subsection{Coarsening to avoid comparing ambiguous levels (validity)}

@@

(Outocmes with very many levels -- even up to of the order of 100 --
aren't really a problem for model fitting with modern software.)
